---
title: "Getting Data into R"
description: "Several ways of getting data into R and performing basic inspections"
output: 
  learnr::tutorial:
    progressive: true
    toc: true
    id: "com.example.tutorials.lab3a"
    version: 1
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(tutorial.helpers)
library(gradethis)
library(tidyverse)
library(knitr)
library(praise)
library(utils)
knitr::opts_chunk$set(echo = T, tidy = T)
```



## What you know already

* Using R as a calculator
* How to create a basic  R script with code and comments.
* Types of variables : factors, strings (characters), numeric (double) and integer, logical .
* Data structures: vectors, lists, matrices, data frames
* How to create vectors and lists and index certain elements in them using  `$` and `[` 

**This Tutorial**

* Getting data into R
* Useful functions to inspect data
* How to summarize the data using measures of center and of spread

## Part 1: Getting Data into R



**Main ways:**


* load an R "object" from a package (e.g. the `datasets` package) - we will look at that first
* load an R "object" from a file (e.g. files that end in `.Rds` or `.rds` or `.rda`) -- this is a type of file you generate from inside R, usually to store R objects or entire sessions in a compressed manner; these files are "R-specific"
* in Rstudio, you can also use the "Import Dataset" option from the file menu -- not ideal in terms of reproducibility but can be helpful for beginners.
* reading in data from a flat text file with delimiters like comma (`.csv` files) or tab -- one of the most common types; we will also see this today.
* reading in data from other data file types* 

*You can import many types of data in R. If you want to see which file types are possible to read in from base R, type `read` in the console and then hit tab. This will show you all the various types of read functions for different file types inherent to base R. 

## Loading data from packages



Although we will later learn how to read data from a spreedsheet into R, let's start by simpling "loading" data into R. A package called `datasets` (already installed with base R) contains several datasets, including this one. 

First, check what objects are already in your workspace. You can do that using the command `ls()`:

```{r ex1,  exercise = T, eval = F}

```


Since we just started, there shouldn't be anything there. If you see anything listed in your workspace, you can remove it with the command `remove()` with the name of the object as argument.

**The Tooth Growth Dataset**

To load the tooth growth data set, simply use the `data` command and use the name of the dataset (`"ToothGrowth"`) as argument

```{r ex2, exercise = T}

```


You can see a list of all datasets available through the `datasets` package by typing `datasets::` and pressing tab. This will attempt to auto-complete and will show you a list you can scroll down on with all the datasets available. Try it:

```{r sandbox-1, exercise = T}

```

If any of them seem interesting, you can use `?name_of_dataset` or `help(name_of_dataset)` to pull up its help page.

*Note*: this is *not* how you usually get data into R. This is a convenient way provided by certain packages which contain pre-organized datasets for you (often for teaching or demo purposes). The most common way is to read a text file into R, as we will see soon, in Part 2.

###


After running the command above, it will appear that nothing happened but, in fact, something did happen. Use the `ls()` command to list all objects in your current workspace:

```{r sandbox-2}

```

*Note*: On the top right of your Rstudio view, you should be able to see a brief description of the dataset.


Voilà! You now have an object called `ToothGrowth` in your workspace.

## Inspecting the Data


In this part we will learn how to get a first feel for the data using some handy functions:

* `help` or `?`
* `class`
* `head`
* `tail`
* `str`
* `summary`
* `dim`


First, let's learn a bit about this dataset by using R's built-in help function. Use either `?` or `help()` below to learn a little about the `ToothGrowth` dataset:

```{r tooth, exercise = T}

```


###

Notice the description of the dataset. Which variables compose this dataset? How would you classify them in terms of their types and subtypes?

Now let's look at the data.


## `class` and `str`

**`class()`**
Once you have your dataset saved as an object in your workspace, one useful command is to know what kind of object it is. Recall the function `class` that we used in our last lab. Try that on the object `ToothGrowth`:

```{r ex3, exercise = T, eval = F} 

```


**`str()`**

A function that many take for granted but which is essential when you're first exploring your data is `str()` (structure).

`str()` will give you a compact display of the contents of your R object.

```{r ex4, exercise = T, eval = T}

```
Each line is referencing one column in your `data.frame`.

Notice that:

* column `len` is of class `numeric`. This is the tooth length of each animal.
* column `supp` is of class `Factor` and it has 2 levels: "OJ", "VC"; orange juice and vitamin C, respectively
* column `dose` is of class `numeric`; 0.5, 1, or 2


This is very helpful indeed! It can guide the next steps in your data analysis.

## `dim`, `nrow`, `ncol`

So the summary on the top right tells you your `ToothGrowth` object has 60 observations of 3 variables, which leads us to `180` "things" to look at. You could also figure this out purely in the console by typing `dim()` and using the name of the object as argument:


```{r dim, exercise = T}

```

Can you guess what the output means? 

###

From the `help(dim)` page you can learn what the output means but I will tell you: it is number of rows and number of columns. 

You can also obtain similar results by running `nrow` and `ncol` separately:

```{r dim2, exercise = T}

```

## `head` and `tail`

Let's say you want to just have a look at the beginning, the first parts of this object. You can do that with the function `head()`, which takes as argument the name of the object. First look up the function using `?` or `help`:

```{r ex8, exercise = T, eval = F}

```

###

Now use it on `ToothGrowth`:

```{r, ex9, exercise = T, eval = F}

```

How many lines do you see in the output? Why do you think that's the number of lines you see? Tip: look at the help function (`?head()`) more closely.


Yes, that's the default value for `head(x, n = 6L)`! What if I only wanted to see the top two lines?

```{r ex5, exercise = T, eval = F}
head(x = ToothGrowth, n = 2)
#or
head(x = ToothGrowth, n = 2L)
#both work
head(x = ToothGrowth, 2)
#also works!
head(n = 2, x = ToothGrowth) #also works!
```

Now try this one. Do you think it will work?

```{r ex6, exercise = T, eval = F}
head(2, ToothGrowth) #does not work
```

*Note*: that one can omit naming the arguments x and n and the function still works. This is *_only true if the order of the arguments follows the default order_*, though. If you want to add arguments in a random order you have to tell the function what they are, as we did above.

### 

What if you wanted to see the last lines of your object?

You guessed correctly. The function is `tail` and it works in the same way! 

Print the last 3 lines of the object `ToothGrowth`:

```{r ex7, exercise = T, eval = T}

```



### `summary`

One final and crucial function to get a sense of your data is `summary()`. The name is pretty descriptive but have a look at the help page for `summary()` anyway. 


```{r ex10, exercise = T, eval = F}

```

Tl;dr: `summary()` will produce summaries for each column in a `data.frame`. Let's give it a try:

```{r ex11, exercise = T, eval = F}

```


## In summary (no pun intended!)

Five functions allowed you to get a feel for this dataset:

* `class`
* `head`
* `tail`
* `str`
* `summary`

These are absolutely essential and you will use them all the time!

## Changing data types of columns

For reflection: go back to `?ToothGrowth` and read about the variables. Do you think
it makes sense to treat the variable `dose` as numeric? If something is a number should it always be treated as a numeric variable?

What if we preferred thinking of `dose` in this dataset as different treatments. I.e., seeing `dose` as a categorical variable rather than a numeric one?

We can convert the class type. 

Extract `dose`, modify it, and put it back into your object:

```{r ex12, exercise = T, eval = T}
#first make a copy of the obj

ToothGrowth2<-ToothGrowth
#now get the column dose

temp_dose<-ToothGrowth$dose
class(temp_dose) #check its class. 
```

Notice that temp dose is simply a vector! You can check by asking:

```{r ex12b,  exercise = T, eval = F, exercise.setup = "ex12"}

is.vector(temp_dose)
```


Now transform it into factor using `as.factor()` and store the new vector in `temp_dose_fact`:

```{r ex12c,  exercise = T, eval = F, exercise.setup = "ex12"}
#transform it into factor using `as.factor()` and store the new vector in `temp_dose_fact`
temp_dose_fact<-as.factor(temp_dose)
#now replace dose in ToothGrowth2 with temp_dose_fact

ToothGrowth2$dose<-temp_dose_fact
#run str again:
str(ToothGrowth2)
#run summary again:
summary(ToothGrowth2)
```

Notice that now the summary of `dose` looks different because the variable is a factor (categorical) instead of numeric. Does it make more sense to you to look at it like this? 


###

Recall that a *matrix* is also a collection of elements of the same type (numerical, character, or logical), but they are organized into a fixed number of rows and columns.

Let's use our `ToothGrowth2` dataset to explore matrices. 

Extract the column `len` from `ToothGrowth2` and assign it to an object called `temp_len`:



```{r ex13setup, exercise = F, eval = T, echo = F}
ToothGrowth2<-ToothGrowth
#now get the column dose

temp_dose<-ToothGrowth$dose
temp_dose_fact<-as.factor(temp_dose)
#now replace dose in ToothGrowth2 with temp_dose_fact

ToothGrowth2$dose<-temp_dose_fact
```

```{r ex13, exercise = T, eval = F, exercise.setup="ex13setup"}

```

Check the class of `temp_len` and whether it is a vector:

```{r ex14setup, echo = F}
ToothGrowth2<-ToothGrowth
#now get the column dose

temp_dose<-ToothGrowth$dose
temp_dose_fact<-as.factor(temp_dose)
#now replace dose in ToothGrowth2 with temp_dose_fact

ToothGrowth2$dose<-temp_dose_fact
temp_len<-ToothGrowth2$len
```

```{r ex14, exercise = T, exercise.setup = "ex14setup"}

```

Ok, now let's make a matrix from this numeric vector and call it `temp_len_mat`

```{r ex14b, exercise=T, exercise.setup = "ex14setup"}
temp_len_mat<-matrix(temp_len)
temp_len_mat
```
It looks like this is not a vector anymore because it has rows and columns. You can check this using the `dim` function (dimensions) on both `temp_len` and `temp_len_mat`:

```{r ex14c, exercise=T, exercise.setup = "ex14setup", eval = F}

```


The second case returns `60 1`, and you already know what that means.

Notice that in the first case the return says `NULL`. That's sbecause it was expecting tabular data. Try `length` instead, which will give you the length of the vector:

```{r ex14d, exercise = T, exercise.setup = "ex14setup", eval = F}

```

By default, `matrix()` coerced your vector into a matrix of 1 column. You could also make a matrix that has more columns:

```{r 14e,exercise=T, exercise.setup = "ex14setup", eval = F}
temp_len_mat2<-matrix(temp_len, ncol = 2) #by default it will try to make a matrix with ncol=1 unless you change it.
dim(temp_len_mat2)
temp_len_mat2
```
It may not be immediately clear to you,but R first filled the 1st column and then the second column. This will be easier to grasp if you use a more illustrative set of numbers. Try this:


```{r matex, exercise = T, eval = F}
asimplemat<-matrix(seq(1:60), ncol = 2)
asimplemat
```



### 

Recall that a `list` is an R object which can contain many different types of elements inside it like vectors, functions and even another list inside it. The sky is the limit!

Let's make a list called `my_list` where we put `ToothGrowth` (original data frame), `ToothGrowth2` (our modified data.frame), `temp_len_mat` (a matrix of one column and 60 rows), `temp_len_mat2` (a matrix of two columns and 30 rows). You simply use the function `list`:

```{r list2, eval = F}
my_list <- list(ToothGrowth, ToothGrowth2, temp_len_mat, temp_len_mat2)
```

Now let's have a look at it using `str()`:

```{r str2, eval = F}
str(my_list)
```
Notice that if you scroll up you can see that it correctly states it is a list of 4 elements. Then it runs the `str()` function for each one of those elements!

## Part 2: descriptive statistics and plots


For this part, we will read in some data into R.

A `.csv` file is a very common format for tables of data. For example, you can export an Excel or google sheets spreadsheet into a `.csv` file very easily.

Here we will use a very simple function from base R called `read.csv`.


```{r setupnum, echo=F, eval=T}
titanicdata<-read.csv("https://raw.githubusercontent.com/bitarellolab/Teaching/main/B215/data/titanic.csv", stringsAsFactors = TRUE)
bat.tongues<-read.csv("https://raw.githubusercontent.com/bitarellolab/Teaching/main/B215/data/BatTongues.csv", stringsAsFactors = TRUE)
guppyFatherSonData <-read.csv("https://raw.githubusercontent.com/bitarellolab/Teaching/main/B215/data/chap02e3bGuppyFatherSonAttractiveness.csv")
```

Use `head`, `tail` and `str`, `dim` on the titanic data to check it out

```{r titanic-head, exercise = T, eval = F, exercise.setup = "setupnum"}
# titanic data
head(titanicdata)
tail(titanicdata)
str(titanicdata)
dim(titanicdata)
```

Now you can also run `summary` on the titanic data:


```{r titanic-summ, exercise = T,eval = F, exercise.setup = "setupnum"}


```


## Summary statistics and missing data!

Now look at age by selecting the column `age` in the `titanicdata`:

```{r titanic3, exercise = TRUE, exercise.setup = "setupnum"}
         
```


If you look through the results, you will see that most values in this list are numbers, but some are NAs. These NAs are the people for which we do not have age information.

By the way, the `titanic.csv` file simply has nothing in the places where there is missing data. See for yourself: [link](https://docs.google.com/spreadsheets/d/111Ez-a6GCeUYcprUzKwcTWE3Adux30UHKRxXDL83Luw/edit?usp=sharing).

When R loaded it, it replaced the empty spots with `NA` automatically!

We have already seen how to calculate the mean of a vector of data using `mean()`. Unfortunately, if there are missing data we need to tell R how to deal with it. A (somewhat annoying) quirk of R is that if we try to take the mean of a list of numbers that include missing data, we get an `NA` for the result! But you already know how to get around this. If you forgot, use the `help()` function for `mean`.

Calculate the mean of the `age` column below:


```{r titanic4, exercise = TRUE, exercise.setup = "setupnum"}
                    
```


Many other functions have this `na.rm = T` option, but for those that don't, you will have to do some data cleaning, which we will learn how to do shortly.

What is the range of values in the age column? Use `range` to find out:

```{r titanic4b, exercise = TRUE, exercise.setup = "setupnum"}
                    
```

What about the interquartile range?

```{r titanic4c, exercise = TRUE, exercise.setup = "setupnum"}
                    
```

And the median?

```{r titanic4e, exercise = TRUE, exercise.setup = "setupnum"}
                    
```

So, should you always set `na.rm = T` just in case? Sure. But you will still want to know whether there are any NAs in your data to begin with, because you should know that instead of simply ignoring it.

One way to do that is to use the function `is.na`. Use it on the `age` column of the `titanicdata` dataset:

```{r titanic5, exercise = TRUE, exercise.setup = "setupnum"}
                    
```


This will return a logical vector of same length as `titanicdata$age` which tells you which rows have NA (TRUE) and which don't (FALSE).

This turned out to be a very long vector! How can we simplify this output? 

We could use the function `table` on the output of the command above. This will give us two columns, one containing the number of results that are `TRUE` and another for `FALSE`.

Le'ts try that here:

```{r titanic6, exercise = TRUE, exercise.setup = "setupnum"}
table(is.na(titanicdata$age))                   
```


Now, try calculating the standard deviation and variance of the age data using the R functions `sd` and `var`. 

```{r titanic7, exercise = TRUE, exercise.setup = "setupnum"}
                    
```


If you notice, variance is just SD squared. Or. SD is just variance square rooted.

Other metrics of variation include the mode and the coefficient of variation. There is no standard function in R to calculate either. 

For the mode, you can use `table` to look at all the values in the column age and then sort them by how prevalent they are using the function `sort`:

```{r titanic8, exercise = TRUE, exercise.setup = "setupnum"}

sort(table(titanicdata$age))
```

And, of course, you could plot the ages and inspect them!

Here is a simple plot to assess that:

```{r titanic9, exercise = TRUE, exercise.setup = "setupnum"}
hist(titanicdata$age)
```
**Note:** technically, age is not a continuous variable, but rather a numerical (discrete) one. But given that there are values across the full range of values, a histogram works ok for this.

THis histogram looks fine but we could make it better for visual inspection and detecting the mode by making the bins smaller, say, going from 0 to 72, given the range we saw above:

```{r titanic10, exercise = TRUE, exercise.setup = "setupnum"}
hist(titanicdata$age, breaks = 0:72)
```
Now you can easily spot the mode!


For the coefficient of variation, you can do this yourself,  directly from the definition: $CV=100\times \frac{S_{X}}{\bar{X}}$

```{r manual-cv, exercise=TRUE, exercise.setup = "setupnum"}
100 * sd(titanicdata$age, na.rm = TRUE) / mean(titanicdata$age, na.rm = TRUE) 

```



### Actually removing missing data from your R object

Lets say you are trying to do a statistical test which lacks na. rm capabilities? Well, it's pretty simple. You can use `na.omit`.

```{r ex-naomit-setup, eval = T, echo = F}
titanicdata<-read.csv("https://raw.githubusercontent.com/bitarellolab/Teaching/main/B215/data/titanic.csv", stringsAsFactors = TRUE)
titanicdata2<-na.omit(titanicdata)
```

```{r ex-naomit, echo = T, eval = F, exercise = T, exercise.setup = "ex-naomit-setup"}
titanicdata2<-na.omit(titanicdata)
```

We can confirm that the missing rows have been deleted by comparing the number of rows in `titanicdata2` with titanic data by using `nrow` on both objects:


```{r ex-naomit2, exercise = TRUE, exercise.setup="ex-naomit-setup"}
nrow(titanicdata)-nrow(titanicdata2)
```



But did this function actually remove all the rows with missing data or did it mess up? 

Well, three functions are useful for us to decide. 

`length` returns the number of entries

`which` returns the element positions for each the logical entry is `TRUE`

`is.na`, which returns a logical statement stating whether any entry in the input is `NA`.

Try testing whether there are any NAs in `titanicdata`  (entire data.frame, not any single column) using `is.na` (which, because we know there are, should return TRUE). Use the command `table` around this to simplify the output.

```{r x10, exercise = TRUE, exercise.setup = "setupnum"}
                    
```

And now try the same command on `titanicdata2`:

```{r x11, exercise = TRUE, exercise.setup = "ex-naomit-setup"}
                    
```

Now, since we know the number of NAs in the age column,we can check whether these numbers are the same. (they are!)

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers"), echo = F}
```
